---
title:          "`hu-neuro-pipeline`"
subtitle:       "Workshop exercises"
author:         "Alexander Enge"
date:           "13/04/2022"
institute:      "Neuro Lab @ Humboldt-Universität zu Berlin"
output:
  github_document:
    toc: true
    toc_depth: 3
---

## Exercises

### 1. Try some Python

The `hu-neuro-pipeline` package requires no Python skills and can be used in its full functionality from within R.
If you would still like to try out running some Python code, this section is for you.
If you want to move straight on how to use the pipeline, please move on to Section 2.

#### Exercise 1.1

Use the `print` function to show the string *Hello world*.
Just as in R, Python strings need to go inside quotation marks.

```{python, eval=FALSE}
print(...)
```

#### Exercise 1.2

Let's use a Python package called `pandas` to read some tabular data.
You can find the behavioral log files from two EEG sessions in the `data/log/` sub-directory.
Load one of them using the `read_csv` function from `pandas` and print the result.
Note that Python always uses `=` rather than `<-` for assigning variables.

```{python, eval=FALSE}
import pandas

log = pandas.read_csv(...)
print(...)
```

#### Exercise 1.3

Next, let's use MNE-Python to process some EEG data.
Checkout the documentation of the package (https://mne.tools) to find a function for reading raw EEG data in BrainVision format.
Then we use the `plot()` method of the `raw` data object to display a random segment of 10 seconds.

```{python, eval=FALSE}
import ...

raw <- ... # Use `preload=True` so that the data are actually loaded
raw.plot(duration=10.0, start=420.0)
```

#### Exercise 1.4

That looks pretty ugly!
Let's apply some filters to clean the data.
Similar to the `plot()` method that we've used above, there is also a method that filters the raw data and can be applied in the same way.
Try to find it in the documentation or try to guess its name and input arguments.
A band-pass filter with a lower edge of 0.1 Hz and an upper edge of 40 Hz should do the trick.
Then we again plot the data so the effect of the filter.

```{python, eval=FALSE}
...
raw.plot(duration=10.0, start=420.0)
```

### 2. Try the pipeline

In this section, you can try out calling the `hu-neuro-pipeline` package from R.

#### Exercise 2.1

Fill in this skeleton for using the `group_pipeline()` function to analyze the UCAP data.
You need to specify the directories where the raw EEG data and log files are stored.
The output directory and EEG trigger values are already provided for you, but you will need to select an ERP component of interest.
Let's call this component `"P1"` and use a time window of 100--150 ms with a region of interest including channels PO3, PO4, POz, O1, O2, and Oz.
The additional `average_by` argument specifies which columns from the log files are used to create by-participant averages for plotting.

```{r, eval=FALSE}
pipeline <- reticulate::import("pipeline")

res <- pipeline$group_pipeline(
  vhdr_files = ...,
  log_files = ...,
  output_dir = "output",
  triggers = c(201:208, 211:218),
  components = list(
    "name" = list(...),
    "tmin" = list(...),
    "tmax" = list(...),
    "roi" = list(...)
  ),
  average_by = c("n_b", "n_b/DeviantPosRL")
)
```

#### Exercise 2.2

Take a look at each of the files that were created in the output directory.
Load the single trial data frame and plot the P1 amplitudes (on the y axis) as a function of the blurr vs. normal condition (on the x axis).

```{r, eval=FALSE}
trials <- readr::read_csv(...)
ggplot(trials, aes(x = ..., y = ...)) +
  geom_dot(position = "jitterdodge")
```

## Solutions

### 1. Try some Python

#### Exercise 1.1

```{python, eval=TRUE}
print("Hello world")
```

#### Exercise 1.2

```{python, eval=TRUE}
import pandas

log = pandas.read_csv("data/log/5_test.txt")
print(log)
```

#### Exercise 1.3

```{python, eval=TRUE}
import mne

raw = mne.io.read_raw_brainvision("data/raw/05.vhdr", preload=True)
raw.plot(duration=10.0, start=420.0)
```

#### Exercise 1.4

```{python, eval=TRUE}
raw.filter(0.1, 40.0)
raw.plot(duration=10.0, start=420.0)
```

### 2. Try the pipeline

#### Exercise 2.1

```{r, eval=FALSE}
library(reticulate)

pipeline <- import("pipeline")

res <- pipeline$group_pipeline(
  vhdr_files = "data/raw",
  log_files = "data/log",
  output_dir = "output",
  triggers = c(201:208, 211:218),
  components = list(
    "name" = list("P1"),
    "tmin" = list(0.1),
    "tmax" = list(0.15),
    "roi" = list(c("PO3", "PO4", "POz", "O1", "O2", "Oz"))
  ),
  average_by = c("n_b", "n_b/DeviantPosRL")
)
```

#### Exercise 2.2

```{r, eval=TRUE}
library(readr)
library(ggplot2)

trials <- read_csv("output/trials.csv")
ggplot(trials, aes(x = n_b, y = P1)) +
  geom_point(position = "jitter")
```
